Blame and Coercion: Together Again for the Second Time

### 2019-10-04 paper.tex
intro

\newtheorem{theorem}{Theorem}[]
\newtheorem{proposition}{Proposition}[]
\newtheorem{corollary}{Corollary}[]
\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\stxrule}[3]{#1 & ::= & #3 & \text{#2}\\}
\newcommand{\stxrulecont}[1]{& | & #1 & \\}
\newcommand{\funrule}[3]{#1 & = & #2 & #3\\}
\newcommand{\comprule}[4]{#1 & \fatsemi^l & #2 & = & #3 & #4 \\}
\newcommand{\plus}[0]{+}
\newcommand{\judgetype}[3]{#1 \vdash #2 : #3}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
	Gradual Typing is great. Implementing gradually typed with blame tracking 
	and space efficiency is tricky. There exist two technique to do this: 
	coercion and threesome. Coercion is easy to understand, and easy enough to 
	implement, but difficult to reason about formally. Threesome is hard to 
	understand, easy to implement, and easy to reason about formally. We 
	propose hyper-coercion, which is easy to understand, as easy to implement 
	as coercion, and easy to reason about.
\end{abstract}

@@

\todo[inline]{KC: I want to add more citations here}

\citet{wadler2009well} introduces blame calculus, an intermediate language for 
gradually typed languages. Blame calculus includes blame labels, which tells 
the position of the cast that fails at run time (blame tracking).

Implementing gradually typed languages on top of blame calculus suffices sever 
space leak. Translation to blame calculus might wrap a tail call with a 
cast. Thus, at runtime, mutually tail-recursive functions can accumulate casts 
at tail position.

In 2007, \citet{herman2010space} proposes a solution to this problem by 
translating Blame Calculus to a variant of \citet{henglein1994dynamic}'s 
Coercions Calculus. The key idea is to represent casts with coercions. 
Coercions can be composed and normalized. Their solution, however, doesn't 
include blame tracking.

@@

\citet{siek2009exploring} incorporate blame tracking into Coercion Calculus by 
decorating coercions with labels. They also propose that there are four blame 
strategies for their Coercion Calculi:
$ \{Lazy, Eager\} \times \{D, UD\} $. 
Lazy strategies blame fewer programs than eager ones,
but also detect less potential type errors.
$ D $ and $ UD $ assign blame labels differently.
They prove that $ Lazy UD $ simulates the Blame Calculus in 
\cite{wadler2009well}. But the counterparts of other strategies in Blame 
Calculus is unknown.

\citet{siek2010threesomes} proposes another approach to combine blame 
tracking and space efficiency. Their solution is based on Threesome Calculus, 
an novel alternative to Coercion Calculus. The key idea is to represent casts 
with threesomes. Threesomes, like coercions, can be composed. There is no 
separate normalization for threesome because every threesome is normalized. 
They prove that their Threesome Calculus bisimulate 
\citet{siek2009exploring}'s Coercion Calculus and
\citet{wadler2009well}'s Blame Calculus.

\citet{siek2012interpretations} introduces a $ lazy D $  Blame Calculus. They 
conjecture that this calculus bi-simulate the $ lazy D $ Coercion Calculi.

Following \citet{siek2010threesomes}, \citet{garcia2013calculating} 
implement all other blame strategies with Threesome Calculus. 
He claims that 
coercion with labels is easy to understand but hard to implement, 
and that 
threesome with labels, however, is easy to implement but hard to understand.
His claim is later affirmed by the group of people who develop threesome 
\citet{siek2015blame}. 
The connection 
between his Threesome Calculi and \citet{siek2009exploring}'s Coercion 
Calculi are established by the fact that the former are derived from the latter.
The connection between these calculi and Blame Calculus, however, 
is still unclear.

\citet{siek2015blame} revisit the coercion-based approach. They simplify the 
Coercion Calculi by only working with coercions in canonical forms.
Fortunately, coercions are canonical when initially constructed, and 
their empowered compose function produces canonical coercions as well. Again 
(\citet{siek2010threesomes}), they delegate all strategies other than 
$ Lazy UD $.

Last year, \citet{kuhlenschmidt2018efficient} present Grift, a space-efficient 
and blame-tracking compiler for a gradually typed language of the same name. 
This implementation is based on the $ Lazy D $ Coercion Calculus.
Their result suggests that implementing coercion is practical. And this is the 
first time product types is considered. Their treatment to products in 
coercions, however, is shown incorrect (\citet{Gradual-TypingType-based-casts}).

Recently, \citet{new2019gradual} show that $ Eager $ strategies are 
incompatible with $\eta$-equivalence of functions, which suggest that these 
strategies are not very ideal. Their research kills questions about $ Eager $ 
strategies, but some questions about $ Lazy $ strategies remain open:

\begin{itemize}
	\item Is the $ Lazy D $ coercion an ideal cast representation?
	\item Does the $ Lazy D $ Coercion Calculus bi-simulate the Blame Calculus?
\end{itemize}

The $ Lazy D $ coercion is claimed easy to understand 
(\cite{garcia2013calculating}, \cite{siek2015blame}) and is shown 
easy enough to implement in a compiler (\cite{kuhlenschmidt2018efficient}).
It is still undesirable, however, in a 
few aspects. Firstly, its compose function is not structurally recursive. Many 
developers of Grift report that it is tricky to convince their 
proof assistants that composition terminates. Secondly, canonical coercion
obscures the nature of a ``normalized cast'' because its definition lies on 
top of coercion.

Perhaps unsurprisingly, threesome has a straightforward recursive 
implementation. And its definition is self-standing. Together with the research 
by \citet{garcia2013calculating}, they suggest that there can be a cast 
representation whose definition is self-standing, and whose composition is 
structurally recursive. Super-coercion introduced in 
\citet{garcia2013calculating} could 
be a promising candidate. However, its definition is a bit complicated. It uses 
10 constructors to deal with an elementary type system with only base types and 
function types. And four constructors are directly related to 
function types. Thus, super-coercion might not scale very well to more 
sophisticated type systems.

We present yet another cast representation, hyper-coercion. Its 
composition is structurally recursive. And its definition is self-standing. 
Besides, there is a clear connection between it and the canonical coercion, 
so we are optimistic that implementing hyper-coercion should be as easy (or 
as hard) as coercion.

@@

We prove formally that $ Lazy D $ (resp. $ Lazy UD $) hyper-coercion calculi 
bi-simulate the $ Lazy D $ (resp. $ Lazy UD $) Blame Calculi. This is possibly 
the first bi-simulation proof for the $ Lazy D $ Blame Calculus.

The structure of this paper is as follows. Section \ref{sec:blame-calculus} 
reviews the state-of-art of $Lazy$ Coercion Calculi. In section 
\ref{sec:hyper-coercion} we present Hyper-coercion. Section \ref{sec:conclude} 
concludes.

@@

$ S \sim T $ reads $ S $ and $ T $ are consistent. Two types are consistent if 
one of them is $ \star $, or they have the same top-most type constructor and 
the corresponding sub-parts are consistent. Consistency is reflexive and 
symmetric, but not transitive.

$ S \smile T $ reads $ S $ and $ T $ are shallowly-consistent. Two types are 
shallowly-consistent if one of them is $ \star $, or they have the same 
top-most type constructor. 
Shallow-consistency is also reflexive, symmetric, and not transitive.

@@

The syntax of $ Lazy D $ Hyper-coercion is shown in \figref{fig:HC-D}.

\begin{theorem}[$ Lazy D $ Hyper-coercion is a proper cast representation]
	\  \\
\begin{enumerate}
	\item If $ v : T $, then $ id(T) \; v = \mathtt{succ} \; v $
	\item If $ v : T_1 $,
	$ c_1 $ is a hyper-coercion from $ T_1 $ to $ T_2 $, and 
	$ c_2 $ is from $ T_2 $ to $ T_3 $,\\
	then $ seq(c_1,c_2) \; v = (c_1 \; v) >>= c_2 $
	\item If $ v : T_1 $ and $ \neg T_1 \smile T_2 $,
	then $ cast(T_1, l, T_2) \; v = \mathtt{fail} \; l $
	\item If $ v : \star $, 
	then $ cast(\star,l,\star) \; v = \mathtt{succ} \; v $
	\item If $ v : P $,
	then $ cast(\star,l,Q) \; (inj \; P \; v) = cast(P,l,Q) \; v $
	\item If $ v : P $,
	then $ cast(P,l,\star) \; v = \mathtt{succ} \; (inj \; P \; v) $
	\item If $ v : \iota $,
	then $ cast(\iota,l,\iota) \; v = \mathtt{succ} \; v $
	\item $ cast(S_1 \rightarrow T_1,l,S_2 \rightarrow T_2) \; (fun \; c_1 \; 
	E \; b \; c_2) $ \\
	$ = \mathtt{succ} \; fun \; seq(cast(S_2,l,S_1),c_1) \; E \; b 
	\; seq(c_2,cast(T_1,l,T_2)) $ 
	\item $ cast(S_1 \times T_1,l,S_2 \times T_2) \; (cons \; v_1 \; c_1 \; 
	v_2 \; c_2) $ \\
	$ = \mathtt{succ} \; (cons \; v_1 \; seq(c_1,cast(S_1,l,S_2)) \; v_2 \; 
	seq(c_2,cast(T_1,l,T_2))) $ 
	\item $ cast(S_1 \plus T_1,l,S_2 \plus T_2) \; (inl \; v \; c) = 
	\mathtt{succ} \; (inl \; v \; seq(c,cast(S_1,l,T_1)) ) $ 
	\item $ cast(S_1 \plus T_1,l,S_2 \plus T_2) \; (inr \; v \; c) = 
	\mathtt{succ} \; (inr \; v \; seq(c,cast(S_2,l,T_2)) ) $ 
\end{enumerate}
\end{theorem}

@@

\begin{corollary}[$ Lazy D $ hyper-coercion is correct]
	If $ \judgetype{\emptyset}{e}{T} $ and $ o : T $
	\[
	e \Downarrow^{D}_{B} o \; \text{if and only if} \; 
	e \Downarrow^{D}_{H} o
	\]
\end{corollary}

>>>>

\newtheorem{theorem}{Theorem}[]
\newtheorem{proposition}{Proposition}[]
\newtheorem{corollary}{Corollary}[]
\newcommand{\figref}[1]{Fig.~\ref{#1}}
\newcommand{\stxrule}[3]{#1 & ::= & #3 & \text{#2}\\}
\newcommand{\stxrulecont}[1]{& | & #1 & \\}
\newcommand{\funrule}[3]{#1 & = & #2 & #3\\}
\newcommand{\comprule}[4]{#1 & \fatsemi^l & #2 & = & #3 & #4 \\}
\newcommand{\plus}[0]{+}
\newcommand{\judgetype}[3]{#1 \vdash #2 : #3}
\newcommand{\lazyUD}{$\mathtt{Lazy UD}$}
\newcommand{\lazyD}{$\mathtt{Lazy D}$}
\newcommand{\eager}{$\mathtt{Eager}$}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
	Gradual Typing is great. Implementing gradually typed with blame tracking 
	and space efficiency is tricky. There exist two technique to do this: 
	coercion and threesome. Coercion is easy to understand, and easy enough to 
	implement, but difficult to reason about formally. Threesome is hard to 
	understand, easy to implement, and easy to reason about formally. We 
	propose hyper-coercion, which is easy to understand, as easy to implement 
	as coercion, and easy to reason about formally.
\end{abstract}

@@

Around 2006, many work on integrating dynamic typing and static typing emerge, 
notably gradual types \cite{siek2006gradual} and hybrid types 
\cite{flanagan2006hybrid}. 
Later \citet{wadler2009well} introduces Blame Calculus, an intermediate 
language for gradual types and hybrid types. 
Blame Calculus throws a blame label when a runtime type-check fails.
Blame labels can include useful information on the failed cast, for example, 
its location in the source code.

Implementing gradually typed languages naively on top of blame calculus 
suffices sever space leak.
Translation to Blame Calculus can wrap some expressions with a cast, including 
those at tail position. Thus, at runtime, mutually tail-recursive functions in 
source language can accumulate casts at tail position.

In 2007, \citet{herman2010space} proposes a solution to this problem by 
translating Blame Calculus to a variant of \citet{henglein1994dynamic}'s 
Coercion Calculus. Their key idea is to represent casts with coercions, 
which can be composed and normalized. Every coercion looks like a list of 
trees. For normalized coercion, the length of the list is bounded by a small 
constant. And the depths of trees do not growth in composition and 
normalization. These two facts implies that by representing casts with 
coercions, we can restore the space-efficiency of tail-recursive programs. This 
Coercion Calculus, however, doesn't include blame tracking.

@@

\citet{siek2009exploring} incorporate blame tracking into Coercion Calculus by 
decorating coercions with labels. They also propose that there are four blame 
strategies for their Coercion Calculi:
$ \{Lazy, Eager\} \times \{D, UD\} $. 
Eager strategies detect more potential type errors, 
but also blame more programs.
$ D $ and $ UD $ assign blame labels differently.
Blame strategy is essentially the semantics of casts.
They prove that the \lazyUD{} Coercion Calculus simulates the Blame Calculus 
in \cite{wadler2009well}. But we don't know whether the simulation also works 
the other way, and what are the Blame counterparts of other Coercion Calculi.

\citet{siek2010threesomes} proposes another approach to combine blame 
tracking and space efficiency. Their solution is based on Threesome Calculus. 
Their key idea is to represent casts with threesomes, a novel cast 
representation. Threesomes, like coercions, can be composed. There is no 
separate normalization because every threesome is in normal form. 
They prove that their Threesome Calculus bi-simulates
the \lazyUD{} Coercion Calculus and the Blame Calculus.

\citet{siek2012interpretations} introduces a \lazyD{} Blame Calculus. They 
conjecture that it bi-simulates the \lazyD{} Coercion Calculus.

Following \citet{siek2010threesomes}, \citet{garcia2013calculating} 
implement all other blame strategies with Threesome Calculus. 
He claims that 
coercion with labels is easy to understand but hard to implement, 
and that 
threesome with labels, however, is easy to implement but hard to understand.
His claim is later affirmed by the group of people who develop threesome 
(\citet{siek2015blame}).
The connection 
between his Threesome Calculi and \citet{siek2009exploring}'s Coercion 
Calculi are established by the fact that the former are derived from the latter.
The connection between eager calculi and Blame Calculus, however, 
is still unclear.

\citet{siek2015blame} revisit the coercion-based approach. They simplify the 
\lazyUD{} Coercion Calculus by only working with normalized coercions.
Fortunately, coercions are trivially normalized when initially constructed, and 
their empowered compose function produces normalized coercions as well. Again 
\cite{siek2010threesomes}, they delegate all other blame strategies.

Last year, \citet{kuhlenschmidt2018efficient} present Grift, a space-efficient 
and blame-tracking compiler for a gradually typed language of the same name. 
This implementation is based on the \lazyD{} Coercion Calculus.
Their result suggests that implementing coercion is practical. They implement 
tuples, which is the first time gradual product types is considered. Their 
treatment on tuples, however, is shown incorrect 
\cite{Gradual-TypingType-based-casts}.

Recently, \citet{new2019gradual} show that \eager{} strategies are 
incompatible with $\eta$-equivalence of functions, which suggest that these 
strategies are not very ideal.

So far, it seems that the coercion-based approach is the best way to combine 
space-efficiency and blame-tracking: coercion is claimed easy to 
understand \cite{garcia2013calculating}\cite{siek2015blame} and is shown 
easy enough to implement in a compiler \cite{kuhlenschmidt2018efficient}.
Coercion is not satisfactory, however, in at least two aspects.
Firstly, its compose function is not structurally recursive, which makes 
it difficult to study coercion formally.
Many developers of Grift report that it is tricky to convince their 
proof assistants that composition terminates. 
Secondly, it is desirable to understand the nature of ``normalized cast'', but 
the coercion-based approach capture this concept indirectly because normalized 
coercion is defined taking a subset of coercion.

Perhaps unsurprisingly, threesome has a straightforward recursive 
implementation. And its definition is self-standing. Together with 
\citet{garcia2013calculating}'s work, they suggest that there can be a cast 
representation whose definition is self-standing, and whose composition is 
structurally recursive. Super-coercion introduced in 
\citet{garcia2013calculating} could 
be a promising candidate. However, its definition is a bit complicated. It uses 
10 constructors to deal with an elementary type system with only base types and 
function types. And four constructors are directly related to 
function types. Thus, super-coercion might not scale very well to more 
sophisticated type systems.

Now we present one such cast representation: hyper-coercion. As we will show in 
Section \ref{sec:hyper-coercion}, there is a clear connection between it and 
the normalized coercion, so we are optimistic that understanding and 
implementing hyper-coercion should be as easy (or as hard).

@@

We prove formally that the \lazyD{} (resp. \lazyUD{}) hyper-coercion calculus 
bi-simulate the \lazyD{} (resp. \lazyUD{}) Blame Calculus. This is possibly 
the first bi-simulation proof for the \lazyD{} Blame Calculus.

The structure of this paper is as follows. 
Section \ref{sec:blame-calculus} reviews the state-of-art of $Lazy$ Blame 
Calculi. 
In section \ref{sec:hyper-coercion} we present Hyper-coercion.
Section \ref{sec:conclude} concludes.

@@

$ S \sim T $ reads ``$ S $ and $ T $ are consistent''.
Two types are consistent if one of them is $ \star $, or they have the same 
top-most type constructor and the corresponding sub-parts are consistent. 
The intuition of $ S \sim T $ is that $ S $ and $ T $ have no conflict type 
information. Consistency is reflexive and symmetric, but not transitive.

One might be tempted to conclude that inconsistency is the root of all runtime 
type errors. In the setting of gradual typing, however, runtime type-checking 
is usually gradual as well. 
$ S \smile T $ reads ``$ S $ and $ T $ are shallowly-consistent''. Two types 
are shallowly consistent if one of them is $ \star $, or they have the same 
top-most type constructor. Shallow-inconsistency is the root of all runtime 
type error -- Casting a value to a shallowly inconsistent type would 
lead to an runtime type error immediately. Shallow-consistency is also 
reflexive, symmetric, and not transitive.

@@

The syntax of \lazyD{} Hyper-coercion is shown in \figref{fig:HC-D}.

\begin{theorem}[\lazyD{} Hyper-coercion is a proper cast representation]
	\  \\
\begin{enumerate}
	\item If $ v : T $, then $ id(T) \; v = \mathtt{succ} \; v $
	\item If $ v : T_1 $,
	$ c_1 $ is a hyper-coercion from $ T_1 $ to $ T_2 $, and 
	$ c_2 $ is from $ T_2 $ to $ T_3 $,\\
	then $ seq(c_1,c_2) \; v = (c_1 \; v) >>= c_2 $
	\item If $ v : T_1 $ and $ \neg T_1 \smile T_2 $,
	then $ cast(T_1, l, T_2) \; v = \mathtt{fail} \; l $
	\item If $ v : \star $, 
	then $ cast(\star,l,\star) \; v = \mathtt{succ} \; v $
	\item If $ v : P $,
	then $ cast(\star,l,Q) \; (inj \; P \; v) = cast(P,l,Q) \; v $
	\item If $ v : P $,
	then $ cast(P,l,\star) \; v = \mathtt{succ} \; (inj \; P \; v) $
	\item If $ v : \iota $,
	then $ cast(\iota,l,\iota) \; v = \mathtt{succ} \; v $
	\item $ cast(S_1 \rightarrow T_1,l,S_2 \rightarrow T_2) \; (fun \; c_1 \; 
	E \; b \; c_2) $ \\
	$ = \mathtt{succ} \; fun \; seq(cast(S_2,l,S_1),c_1) \; E \; b 
	\; seq(c_2,cast(T_1,l,T_2)) $ 
	\item $ cast(S_1 \times T_1,l,S_2 \times T_2) \; (cons \; v_1 \; c_1 \; 
	v_2 \; c_2) $ \\
	$ = \mathtt{succ} \; (cons \; v_1 \; seq(c_1,cast(S_1,l,S_2)) \; v_2 \; 
	seq(c_2,cast(T_1,l,T_2))) $ 
	\item $ cast(S_1 \plus T_1,l,S_2 \plus T_2) \; (inl \; v \; c) = 
	\mathtt{succ} \; (inl \; v \; seq(c,cast(S_1,l,T_1)) ) $ 
	\item $ cast(S_1 \plus T_1,l,S_2 \plus T_2) \; (inr \; v \; c) = 
	\mathtt{succ} \; (inr \; v \; seq(c,cast(S_2,l,T_2)) ) $ 
\end{enumerate}
\end{theorem}

@@

\begin{corollary}[\lazyD{} hyper-coercion is correct]
	If $ \judgetype{\emptyset}{e}{T} $ and $ o : T $
	\[
	e \Downarrow^{D}_{B} o \; \text{if and only if} \; 
	e \Downarrow^{D}_{H} o
	\]
\end{corollary}
